---
title: "Financial Dependence and Healthcare Utilization among Older Adults"
subtitle: "Evidence from the Longitudinal Ageing Study in India (LASI) Wave 1, 2017–18"
author:
  - name: Ramesh Beniwal
    affiliations:
      - name: Azim Premji University, Bhopal
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    transition: fade
    background-transition: fade
    incremental: false
    width: 1280
    height: 720
    margin: 0.06
    fig-align: center
    echo: false
    embed-resources: true
    footer: "Student Project Conclave 2026 | Azim Premji University, Bhopal"
    title-slide-attributes:
      data-background-color: "#0D2240"
  pptx:
    slide-level: 2
execute:
  warning: false
  message: false
  echo: false
editor:
  markdown:
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

library(haven)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stargazer)

HH_FI <- read_dta("HH_FI.dta")

HH_FI <- HH_FI |>
  mutate(across(where(haven::is.labelled), as.numeric)) |>
  mutate(
    gender_label    = factor(sex, levels = c(0, 1), labels = c("Female", "Male")),
    zone_label      = factor(zone),
    residence_label = factor(residence_type, levels = c(0, 1), labels = c("Urban", "Rural")),
    class_caste     = case_when(
      caste_social_group == 4             ~ "Upper Caste",
      caste_social_group %in% c(1, 2, 3) ~ "Lower Caste",
      TRUE                                ~ NA_character_
    )
  )

pub_theme <- theme_minimal(base_size = 14) +
  theme(
    plot.title       = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 8)),
    plot.subtitle    = element_text(size = 12, hjust = 0.5, color = "grey40", margin = margin(b = 10)),
    plot.caption     = element_text(size = 9, color = "grey55", hjust = 0, margin = margin(t = 10)),
    axis.title       = element_text(size = 13, face = "bold"),
    axis.text        = element_text(size = 12),
    legend.title     = element_text(size = 13, face = "bold"),
    legend.text      = element_text(size = 12),
    legend.position  = "bottom",
    strip.text       = element_text(size = 13, face = "bold"),
    strip.background = element_rect(fill = "#f0f4f8", color = "grey75"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    panel.spacing    = unit(1.5, "lines"),
    plot.margin      = margin(10, 10, 10, 10)
  )
theme_set(pub_theme)

gender_colors    <- c("Female" = "#e05c5c", "Male" = "#3a7abf")
insurance_colors <- c("No Insurance" = "#e05c5c", "Insured" = "#3a7abf")
pension_colors   <- c("No Pension" = "#e09a3a", "Pension" = "#3a7abf")
```


## Motivation 

:::: {.columns}
::: {.column width="50%"}
- India is undergoing a rapid demographic transition with a growing elderly population.(**45+ age**)
- **Globally:**13.9% (1.1 billion) older persons in 2022. By 2050, expected to double to 2.1 billion.
- **India:** 10.5% in 2021. By 2050, the share of older persons will double to 20.8% (347 million) (**UNFPA**)
- **Population ageing is associated with:**
- Higher prevalence of chronic diseases and disabilities
- Increased demand for healthcare services and long-term care
:::

::: {.column width="50%"}
```{r}
#| label: fig
#| fig-width: 10
#| fig-height: 5
#| fig-align: center
library(tidyverse)

# Create tibble with NFHS data
older_persons_india <- tibble(
  year = c(1992, 1998, 2005, 2015, 2021),
  percentage = c(15.43, 15.88, 17.78, 22.04, 24.85)
)

older_persons_india %>%
  ggplot(aes(x = year, y = percentage)) +
  geom_line(size = 1.2, color = "#1f77b4") +
  geom_point(size = 3, color = "#1f77b4") +
  scale_x_continuous(breaks = c(1992, 1998, 2005, 2015, 2021)) +
  labs(
    title = "Percentage of Older Persons in India (1992-2021)",
    subtitle = "NFHS Rounds 1, 2, 3, 4, and 5",
    x = "Year",
    y = "Percentage (%)",
    caption = "Source: NFHS 1, 2, 3, 4, 5"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "grey40"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 11)
  )
```
:::
::::


## Motivation Contd...

:::: {.columns}
::: {.column width="100%"}

- **Traditional family-based support systems are weakening:**
- Decline in joint family structures and increasing migration
- Rising risk of financial and social vulnerability among the elderly
- **Healthcare needs of the elderly are high but utilization remains unequal:**
- Access to healthcare is strongly influenced by socioeconomic status
- **Financial constraints are a key barrier:**
- Limited evidence on the combined role of insurance, pensions, and financial support
- **Contribution of this paper:**
- Examines how financial security affects healthcare-seeking behavior among the elderly
- Provides policy insights for ageing and social protection in developing countries
:::
::::



## Data Source: LASI Wave 1, 2017–18 
:::: {.columns}

::: {.column width="60%"}
**Longitudinal Ageing Study in India (LASI)**
- So this study uses the Longitudinal Ageing Study in India, Wave 1, conducted in 2017-18, NFHS(National Family Health Survey) and UNFPA (United Nations Population Fund). We used individual-level and household-level dataset from LASI for this study. This study has covered adults aged 45 to 100 across all Indian states and union territories.

- LASI-World largest dataset on elderly population. 



**Dependent Variables**

| Variable | Description |
|----------|-------------|
| Healthcare | Any facility visit (past 12 months) |
| Inpatient Hospitalization | Any hospital admission |

:::

::: {.column width="40%"}
**Independent Variables**

- Health Insurance Coverage
- Pension Status
- Financial Support from family/others
- Monthly Per-capita Consumer Expenditure (MPCE) Quintiles

**Controls**

- Age, Sex, Residence type
- Religion, Caste
- **State Fixed Effects**
:::

::::


## Figure 1: Healthcare Visit by MPCE Quintile - Descriptive Analysis 

```{r}
#| label: fig1-healthcare-mpce
#| fig-width: 10
#| fig-height: 5
#| fig-align: center

HH_FI |>
  filter(!is.na(mpce_quintile), !is.na(healthcare)) |>
  mutate(MPCE = factor(mpce_quintile, levels = 1:5,
                       labels = c("Poorest", "Poor", "Middle", "Rich", "Richest"))) |>
  group_by(MPCE) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = MPCE, y = Visit, fill = MPCE)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")), vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_brewer(palette = "Blues") +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Healthcare Visit by Economic Status (MPCE Quintile)",
       x = "MPCE Quintile", y = "% Visited Any Healthcare Facility",
       caption = "Source: LASI Wave 1, 2017–18.") +
  theme(legend.position = "none")
```

## Figure 2: Healthcare Visit by Caste Class- Descriptive Analysis

```{r}
#| label: fig2-healthcare-caste
#| fig-width: 8
#| fig-height: 5
#| fig-align: center

HH_FI |>
  filter(!is.na(class_caste), !is.na(healthcare)) |>
  mutate(Caste = factor(class_caste, levels = c("Lower Caste", "Upper Caste"))) |>
  group_by(Caste) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = Caste, y = Visit, fill = Caste)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")), vjust = -0.5, size = 6, fontface = "bold") +
  scale_fill_manual(values = c("Lower Caste" = "#E69F00", "Upper Caste" = "#56B4E9")) +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Healthcare Visit by Caste Class",
       x = "Caste Class", y = "% Visited Any Healthcare Facility",
       caption = "Source: LASI Wave 1, 2017–18.") +
  theme(legend.position = "none")
```

## Figure 3: Healthcare Visit by MPCE Quintile & Caste- Descriptive Analysis

```{r}
#| label: fig3-healthcare-mpce-caste
#| fig-width: 11
#| fig-height: 5
#| fig-align: center

HH_FI |>
  filter(!is.na(class_caste), !is.na(mpce_quintile), !is.na(healthcare)) |>
  mutate(
    Caste = factor(class_caste, levels = c("Lower Caste", "Upper Caste")),
    MPCE  = factor(mpce_quintile, levels = 1:5,
                   labels = c("Poorest", "Poor", "Middle", "Rich", "Richest"))
  ) |>
  group_by(MPCE, Caste) |>
  summarise(Visit = mean(healthcare, na.rm = TRUE) * 100, .groups = "drop") |>
  ggplot(aes(x = MPCE, y = Visit, fill = Caste)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = paste0(round(Visit, 1), "%")),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Lower Caste" = "#E69F00", "Upper Caste" = "#56B4E9")) +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Healthcare Visit by Economic Status and Caste Class",
       x = "MPCE Quintile", y = "% Visited Any Healthcare Facility", fill = "Caste Class",
       caption = "Source: LASI Wave 1, 2017–18.")
```

## Figure 4: Healthcare Utilization & Financial Protection by Gender - Descriptive Analysis

```{r}
#| label: fig4-descriptive-gender
#| fig-width: 12
#| fig-height: 5
#| fig-align: center

HH_FI |>
  filter(!is.na(sex)) |>
  mutate(Gender = factor(sex, 0:1, c("Female", "Male"))) |>
  group_by(Gender) |>
  summarise(
    `Any Healthcare\nVisit`      = mean(healthcare,               na.rm = TRUE) * 100,
    `Inpatient\nHospitalization` = mean(inpatient_hospitalization, na.rm = TRUE) * 100,
    `Health Insurance\nCoverage` = mean(health_insurance_covered,  na.rm = TRUE) * 100,
    `Pension\nReceipt`           = mean(pension_status,            na.rm = TRUE) * 100,
    `Financial\nSupport`         = mean(financial_support,         na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  pivot_longer(cols = -Gender, names_to = "Measure", values_to = "Value") |>
  mutate(
    Category = case_when(
      Measure %in% c("Any Healthcare\nVisit", "Inpatient\nHospitalization") ~ "Healthcare Utilization",
      TRUE ~ "Financial Protection"
    ),
    Category = factor(Category, levels = c("Healthcare Utilization", "Financial Protection")),
    Measure  = factor(Measure, levels = c(
      "Any Healthcare\nVisit", "Inpatient\nHospitalization",
      "Health Insurance\nCoverage", "Pension\nReceipt", "Financial\nSupport"
    ))
  ) |>
  ggplot(aes(x = Measure, y = Value, fill = Gender)) +
  geom_col(position = position_dodge(width = 0.65), width = 0.55, color = "white", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(Value, 1), "%")),
            position = position_dodge(width = 0.65), vjust = -0.5, size = 3.5, fontface = "bold") +
  facet_wrap(~Category, scales = "free_x", nrow = 1) +
  scale_fill_manual(values = gender_colors) +
  scale_y_continuous(limits = c(0, 90), labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Healthcare Utilization and Financial Protection by Gender",
       x = NULL, y = "Rate (%)", fill = "Gender",
       caption = "Source: LASI Wave 1, 2017–18.")
```

## Regression Analysis {background-color="#0D2240" .center}

[ We need regression to look at how correlation has been established. Now let's talk about numbers, magnitudes of relationships.]{style="color: white;"}

[ By OLS.]{style="color: white;"}

[ The variation of the state - controlled]{style="color: white;"}

## Regression 1: Healthcare Utilization and Insurance Coverage
```{r}
#| label: reg1-insurance
#| results: asis

hc_ins_1 <- lm(healthcare ~ health_insurance_covered + factor(state_id), data = HH_FI)
hc_ins_2 <- lm(healthcare ~ health_insurance_covered +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)
ip_ins_1 <- lm(inpatient_hospitalization ~ health_insurance_covered + factor(state_id), data = HH_FI)
ip_ins_2 <- lm(inpatient_hospitalization ~ health_insurance_covered +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)

stargazer(hc_ins_1, hc_ins_2, ip_ins_1, ip_ins_2,
          type              = "html",
          keep              = "health_insurance_covered",
          omit.stat         = c("f", "ser", "adj.rsq", "rsq"),
          column.labels     = c("Healthcare", "Healthcare", "Inpatient", "Inpatient"),
          dep.var.caption   = "",
          add.lines         = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE",  "Yes", "Yes", "Yes", "Yes")
          ),
          omit.table.layout = "n",
          table.attr        = 'style="width:90%; margin:auto; font-size:0.88em;"',
          header            = FALSE)
```

*\*p<0.1; \*\*p<0.05; \*\*\*p<0.01*

## Regression 2: Healthcare Utilization and Pension Receipt
```{r}
#| label: reg2-pension
#| results: asis

hc_pen_1 <- lm(healthcare ~ pension_status + factor(state_id), data = HH_FI)
hc_pen_2 <- lm(healthcare ~ pension_status +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)
ip_pen_1 <- lm(inpatient_hospitalization ~ pension_status + factor(state_id), data = HH_FI)
ip_pen_2 <- lm(inpatient_hospitalization ~ pension_status +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)

stargazer(hc_pen_1, hc_pen_2, ip_pen_1, ip_pen_2,
          type              = "html",
          keep              = "pension_status",
          omit.stat         = c("f", "ser", "adj.rsq", "rsq"),
          column.labels     = c("Healthcare", "Healthcare", "Inpatient", "Inpatient"),
          dep.var.caption   = "",
          add.lines         = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE",  "Yes", "Yes", "Yes", "Yes")
          ),
          omit.table.layout = "n",
          table.attr        = 'style="width:90%; margin:auto; font-size:0.88em;"',
          header            = FALSE)
```

*\*p<0.1; \*\*p<0.05; \*\*\*p<0.01*

## Regression 3: Healthcare Utilization and Financial Support
```{r}
#| label: reg3-financial
#| results: asis

hc_fin_1 <- lm(healthcare ~ financial_support + factor(state_id), data = HH_FI)
hc_fin_2 <- lm(healthcare ~ financial_support +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)
ip_fin_1 <- lm(inpatient_hospitalization ~ financial_support + factor(state_id), data = HH_FI)
ip_fin_2 <- lm(inpatient_hospitalization ~ financial_support +
                  age + sex + residence_type +
                  factor(religion) + factor(caste_social_group) +
                  factor(marital_status) + factor(state_id), data = HH_FI)

stargazer(hc_fin_1, hc_fin_2, ip_fin_1, ip_fin_2,
          type              = "html",
          keep              = "financial_support",
          omit.stat         = c("f", "ser", "adj.rsq", "rsq"),
          column.labels     = c("Healthcare", "Healthcare", "Inpatient", "Inpatient"),
          dep.var.caption   = "",
          add.lines         = list(
            c("Controls", "No", "Yes", "No", "Yes"),
            c("State FE",  "Yes", "Yes", "Yes", "Yes")
          ),
          omit.table.layout = "n",
          table.attr        = 'style="width:90%; margin:auto; font-size:0.88em;"',
          header            = FALSE)
```

*\*p<0.1; \*\*p<0.05; \*\*\*p<0.01*


## Conclusion

:::: {.columns}

::: {.column width="100%"}
1. Higher MPCE quintile → higher healthcare utilization
2. Caste-based inequalities persist in access
3. Women visit more but have far less formal financial protection
4. Insurance matters most for the **oldest (75+)** and most vulnerable
5. Pension income strongly **enables hospital care**, especially in rural areas
6. Addressing this now is critical — as India's older population grows, gaps in financial protection and healthcare access will create serious instability at both household and national level.
:::

::::


## {.center-middle}

<div class="thankyou">
Thank You
</div>

---

## {.center-middle}

<div class="questions">
Questions & Discussion
</div>

<br><br>

<div class="contact">

**Ramesh Beniwal **  
ramesh.beniwal24_ug@apu.edu.in  

<br>

**Ram Krishna**  
ram.krishna24_ug@apu.edu.in  

</div>





